---
title: 'Table 2 Fallacy'
author: "Brad Hackinen"
date: "Fall 2025"
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(fixest)
```

# The Table 2 Fallacy

It is common for researchers to present a table of estimation results that includes both the main treatment effect and the coefficients on the control variables. This is fine.

However, people often drift into interpreting the coefficients on the control variables as if they were causal effects. This is an error.

The set of variables that we select as control variables generally will not also work as a set of control variables for measuring a different causal effect.

The following example illustrates this point. Given the DAG describing how the data is generated, the backdoor criterion suggests three valid adjustment sets for estimating the causal effect of $D$ on $Y$:

{X}
{W}
{X,W}

All three sets of controls can be used to estimate the ATE (equal to 1 in this demo). However, the coefficient estimates on the control variables can change dramatically depending on which set of controls is used.

```{r}
N <- 100000

data <- tibble(
  X = rnorm(N),
  W = X + rnorm(N),
  D = X + rnorm(N) > 0,
  Y = D + W + rnorm(N)
)

fit1 <- feols(Y ~ D,data)
fit2 <- feols(Y ~ D + X,data)
fit3 <- feols(Y ~ D + W,data)
fit4 <- feols(Y ~ D + X + W,data)

etable(fit1,fit2,fit3,fit4)
```